Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-13 doc/manuscript/ESSDmanuscript.Rmd
dead wood

#Cited sections (listed here because these can change)
section_C_pools = 2.1

@@

Dead organic matter includes all non-living biomass that is not within the mineral soil layer and smaller than the litter size threshold.

Dead wood is defined as...

Litter is defined as including " all  non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil. This includes litter (OL),  fumic (OF),  and  humic (OH) layers.  Live  fine  roots  (of  less  than  the  suggested  diameter  limit for  belowground  biomass) are included in litter where they cannot be distinguished from it empirically." (2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)

Not required for forest land remaining forest land under Tier 1 methodology [@ipcc_2006_2006].

@@

<!--dead wood and litter comparisons will be particularly interesting, as IPCC values are based on just a handful of references for each climate zone (table 2.2 in 2019 guidelines)-->

@@

>>>>

#Cited sections (listed here because these can change)
section_C_pools = 2.1 
section_land_use_categories = 2.2

@@

Dead organic matter includes all non-living biomass that is not within the mineral soil layer and smaller than the litter size threshold. 
It's inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].

Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_2006_2006]. 
This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
While dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance; @carmona_coarse_2002], aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
In turn, they are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

Litter is defined as including " all  non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil. This includes litter (OL),  fumic (OF),  and  humic (OH) layers.  Live  fine  roots  (of  less  than  the  suggested  diameter  limit for  belowground  biomass) are included in litter where they cannot be distinguished from it empirically." (2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->

@@

@@

Given widespread trends of increasing tree mortality[@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.

### 2022-04-13 doc/manuscript/ESSDmanuscript.tex
dead wood

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation & min dbh & acceptable to exclude understory if minor component\\
\hline
 &  & include non-dicot trees? & ?\\
\hline
 &  & include dead standing? & no\\
\hline
 &  & biomass allometry & ?\\
\hline
belowground biomass & all biomass of live roots & all factors relevant to aboveground biomass & see above\\
\hline
 &  & allometry or assumed ratio of below- to above-ground biomass (R) & can estimate based on R\\
\hline
 &  & min root diameter & fine roots may be excluded\\
\hline
 &  & sampling depth & ?\\
\hline
dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & include belowground? & \vphantom{1} yes\\
\hline
litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter (= max diameter for deadwood) & 10 cm default, but may be chosen by country\\
\hline
 &  & layers included & includes entire O horizon\\
\hline
 &  & include belowground? & yes\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & depth = 30 cm default, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

@@

Dead organic matter includes all non-living biomass that is not within
the mineral soil layer and smaller than the litter size threshold.

Dead wood is defined as\ldots{}

Litter is defined as including " all non-living biomass with a diameter
less than a minimum diameter chosen by the country (for example 10 cm),
lying dead, in various states of decomposition above the mineral or
organic soil. This includes litter (OL), fumic (OF), and humic (OH)
layers. Live fine roots (of less than the suggested diameter limit for
belowground biomass) are included in litter where they cannot be
distinguished from it empirically." (2003 IPCc GPG for LULUCF
(https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)

Not required for forest land remaining forest land under Tier 1
methodology \citep{ipcc_2006_2006}.

@@

>>>>

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation & min dbh & may exclude understory if minor component\\
\hline
 &  & include non-dicot trees? & ?\\
\hline
 &  & include dead standing? & no\\
\hline
 &  & biomass allometry & ?\\
\hline
belowground biomass & all biomass of live roots & all factors relevant to aboveground biomass & see above\\
\hline
 &  & allometry or assumed ratio of below- to above-ground biomass (R) & can estimate based on R\\
\hline
 &  & min root diameter & may exclude fine roots\\
\hline
 &  & sampling depth & ?\\
\hline
dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & include belowground? & \vphantom{1} yes\\
\hline
litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter (= max diameter for deadwood) & 10 cm default, but may be chosen by country\\
\hline
 &  & layers included & includes entire O horizon\\
\hline
 &  & include belowground? & yes\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & 30 cm default, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

@@

Dead organic matter includes all non-living biomass that is not within
the mineral soil layer and smaller than the litter size threshold. It's
inclusion in inventories is not required under Tier 1 methodology for
Forest Land remaining Forest Land (see section 2.2), but is required for
land that has transitioned to or from forest within the past 20 years
\citep{ipcc_2006_2006}.

Dead wood, which is typically \textless50 t C ha\textsuperscript{-1} but
can exceed 150 t C ha\textsuperscript{-1}
\citep{anderson-teixeira_carbon_2021}, is defined by IPCC as ``all
non-living woody biomass not contained in the litter, either standing,
lying on the ground, or in the soil'' \citep{ipcc_2006_2006}. This pool
includes standing and fallen dead wood, stumps, and dead roots of
diameter ≥10 cm (or a diameter specified by the country). While dead
wood stocks and fluxes can be quite variable across forests
\citep{anderson-teixeira_carbon_2021}, and can at times be the dominant
pool in a forest ecosystem {[}e.g., following a severe natural
disturbance; \citet{carmona_coarse_2002}{]}, aboveground dead wood
remains relatively poorly characterized at a global scale
\citep{anderson-teixeira_carbon_2021}, and belowground dead wood is
rarely studied \citep{merganicova_dadwood_2012}. In turn, they are
poorly characterized in large-scale forest C budgets
\citep{pan_large_2011, harris_global_2021}, and IPCC's latest Tier 1
default values are based on just 1-31 references per climate zone
\citep[Table 2.2 in][]{ipcc_2019_2019}.

Litter is defined as including " all non-living biomass with a diameter
less than a minimum diameter chosen by the country (for example 10 cm),
lying dead, in various states of decomposition above the mineral or
organic soil. This includes litter (OL), fumic (OF), and humic (OH)
layers. Live fine roots (of less than the suggested diameter limit for
belowground biomass) are included in litter where they cannot be
distinguished from it empirically." (2003 IPCc GPG for LULUCF
(https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
IPCC's latest Tier 1 default values are based on just 1-7 references per
climate zone \citep[Table 2.2 in][]{ipcc_2019_2019}.

@@

Given widespread trends of increasing tree mortality\citep{refs},
including through severe natural disturbance \citep{refs}, better
characterization of dead wood will be critical.
